An Analysis of the Feasibility and Performance of BitTorrent as a Data Transfer Protocol in Peer to Peer Backup Systems

### 2014-04-10 paper/paper.tex
Paper: rework Peer to Peer Networks / Peer to Peer Backup Systems to appeal to a broader audience and feel less like an analysis

\begin{document}
\newpage
\begin{center}
\thispagestyle{empty}
\Large{\textbf{A Major Qualifying Project Report\\ \large{ON}}}\\[0.7cm]
\LARGE{\textsc {\textbf{An Analysis of the Feasibility and Performance of BitTorrent as a Data Transfer Protocol in Peer to Peer Backup Systems}}}\\[0.5cm]
\vspace{0.2cm}
\Large{\textbf{\\Submitted to the Faculty of }}
\LARGE{\textbf{\\WORCESTER POLYTECHNIC INSTITUTE\\}}
\vspace{0.5cm}
\Large{\textbf{\\In Partial Fulfillment of the Requirement for the}}
\Large{\textbf{\\Degree of Bachelor of Science}}
\vspace{0.5cm}
\Large{\textbf{\\by}}\\[0.5cm]
\large{\textbf{Dan Bouffard}}\\
\large{\textbf{Marc Green}}\\
\vspace{0.5cm}
\large{\textbf{UNDER THE GUIDANCE OF}}\\
\large{\textbf{Professor G\'abor N. S\'ark\"ozy}}\\
\vspace{1cm}
\large{\textbf{\\May 1, 2014}}\\
\newpage
\end{center}

@@

\chapter{Background Information}
\section{Centralized Backup Systems}
% TODO: get reliable sources for number of users of OneDrive and Google Drive
Several popular backup systems currently in use have a centralized system design. In these systems, there are many clients that all back up their data to a central location controlled by a single entity. Common elements of these systems include a fixed amount of data storage per person, with the option to obtain more for a fee. Two of the more popular systems are Dropbox, which has over 200 million users \cite{dropboxusers}, OneDrive, with over 250 million users, and Google Drive, with over 120 million users.

@@

However, there are also several problems with using a centralized system to backup data. One of these problems is that there is a single point of failure, should the storage site be compromised in some way (for example, natural disasters). This is less of a problem for big companies who offer the service, as they will often have the resources to handle problems, but they still need to be taken into consideration when creating the system, at a cost to the maintainers.

@@

% TODO: reliable sources for data caps and monthly/yearly fees
Finally, there are often restrictions on what can be stored using the service. This includes both the size of individual files and the amount of total data. More space is gained by paying a monthly or yearly fee, and even then the amount of space is capped at a specific size. For users who need to backup a large amount of large files, these systems are not very useful.

@@

A peer to peer (P2P) network is a type of distributed network model in which participants form direct connections to each other. This is in contrast to the centralized client-server network model, in which all participating clients connect to a central server to carry out their task. In peer to peer networks, each participant, or "peer", functions as both a client and a server; peers initiating a request take on the role of the client, and peers answering the request take on the role of the server.

In the client-server model, servers are expected to be always available and accessible. Peer to peer networks do not have this luxury. Instead, they experience the constant fluctuation of peers joining and leaving the network. This is called \textit{churn}. Churn complicates data storage and retrieval because there is no guarantee that a peer that has some desired data will always be available. The strategies for dealing with churn, and many other P2P-specific complications, depend on the type of overlay network a P2P application uses. We discuss churn in more detail in the \textit{Peer to Peer Backup Systems} section.
% TODO link to Peer to Peer backup systems section instead of just mentioning it

\subsection{Overlay Networks}

% TODO: source this?
At their foundation, peer to peer networks are implemented through overlay networks on top of existing infrastructure, such as the Internet. The responsibilities of a P2P overlay network includes providing peers the ability to join and leave the network and find other peers.

@@

These overlay networks, and how they implement their responsibilities, are discussed below.

% For each overlay, discuss:
% - Joining
% - Finding Peers
% - Situations where each are appropriate
\hfill \\
\textit{Structured Overlay Networks}
\hfill \\

Structured overlay networks are defined by a virtual address space and a topology. For example, peers could be organized in a ring and addressed by consecutive numbers, incrementing counter-clockwise. Or, peers could be organized in a 3D cube and addressed by their (x,y,z) coordinates.
% TODO maybe insert figure of chord ring topology

The purpose of peer to peer networks (more broadly, networks in general) is to share data. The data that is present in a structured overlay network shares the same aforementioned virtual keyspace. That is, structured overlay networks map the data into a valid address through some function. For example, if peers were addressed by a hash of their IP address (which could be organized into a ring), the data could also be hashed to find its "address". This is done to determine which peer is responsible for a given piece of data. Specifically, a peer is responsible for all data in the keyspace that the peer is closest to. Thus, the size of the keyspace for which a peer is responsible decreases as more peers join the network.

% TODO cite router.bittorrent.com as well known bootstrap node?
To join a structured overlay network, a peer must know at least one other peer in the network. This peer is often called the \textit{bootstrap peer}. The joining peer will contact the bootstrap peer to find its proper place in the topology. More details are given below, in the explanation of how to \textit{find other peers}. In BitTorrent's structured overlay network \footnote{To clarify, we are referring to BitTorrent's distributed hash table, not its unstructured network. The latter involves using centralized trackers, the former removes such dependency. We do not yet use the term \textit{distributed hash table} to keep the explanation simple without loss of generality. We discuss distributed hash tables elsewhere in the paper.}, a well known bootstrap node \footnote{The BitTorrent specification uses \textit{node} to refer to participants of its structured overlay network and \textit{peer} as a more general term \cite{bittorrentDHT}. We will adopt this convention when speaking about BitTorrent, and use them interchangeably otherwise.} is router.bittorrent.com. A BitTorrent node randomly chooses a 160 bit \textit{nodeID} to use as an address. It will contact a bootstrap node to find the nodes it is closest to in the topology/address space. Closeness, in this case, is determined by taking the $XOR$ of \textit{nodeID}s; smaller values are defined as closer \cite{bittorrentDHT}.

Leaving a structured overlay network can be as simple as not responding to any messages. Some specifications leave this functionality undefined, perhaps because peers may simply shutdown instead of gracefully quitting. Such overlay networks handle node departures as node failures.

To find other peers, a peer in a structured overlay network will consult its routing table. This table is initially populated upon joining the network, and is kept up to date as peers join and leave. When a joining peer first contacts a bootstrap peer, it is unlikely that the bootstrap peer has the joining peer's direct neighbors in its routing table. This is because the routing table is designed to include many peers that are close, and only a few peers that are far. A joining peer, by asking the bootstrap node, will learn of the peers in the bootstrap peer's routing table that the joining node is close to. The joining peer will then contact these peers, ask for the same information, and learn of peers it is even closer to. This process repeats (usually in $O(\log n)$ time) until the joining node learns its direct neighbors \cite{p2pSurvey}.

Structured overlay networks excel at locating specific information very quickly. This is because their virtual address space provides full accountability of any piece of data. % mention keyword searches?

\hfill \\
\textit{Unstructured Overlay Networks}
\hfill \\

Unstructured overlay networks do not enforce any particular address space or topology on peers. Instead, the networks typically resemble random graphs. There are no standardized techniques for peers in unstructured overlay networks to find other nodes or search for data. Gnutella, for example, requires a peer to contact a bootstrap peer to initially join the network. The joining peer then sends \textit{ping} messages with increasing TTL values to the bootstrap peer, which forwards the ping to its neighbors. This continues until the joining peer has received enough \textit{pong} messages from peers willing to neighbor \cite{p2pSurvey}. % [Discuss an example of searching in unstructured networks -- see survey paper, or maybe random walk paper]

The lack of standard techniques permits a large degree of freedom in these networks. % [Talk more about this]

Unstructured overlay networks excel at finding content that is popular. This is because the search techniques are more likely to find data that exist on many peers than data that exist on fewer. Contrast this to structured overlay networks, which are good for finding unpopular data because it is easy to determine which peer is responsible for it. % mention that strucutred is _bad_ for popular data? why is it?

\hfill \\
\textit{Hierarchical Overlay Networks}
\hfill \\

\section{Peer to Peer Backup Systems}
Peer to Peer Backup Systems are networks that allow users to backup data without the use of a central entity. Instead, each user offers some of his or her storage space in order to use the space of others as backup locations. There are several common components of existing P2P backup systems which are described in the following sections.

\subsection{File Chunking}
The format in which files are stored can have an effect on the efficiency of the system. pStore and PeerStore split files up into chunks, rather than distributing files to peers in their entirety. They also use special files, called File Block Lists in pStore and Descriptors in PeerStore, that list which blocks belong to the file and where to find them in the network. This has the benefit of having data units be a fixed size that will be easy to store for any node. For example, if a peer needs to store a 10GB file, it does not need to find a peer capable of storing 10GB (which might be rare); rather, it needs to find $file\_size/chunk\_size*number\_of\_replicas$ peers. Although lowering the chunk size will increase the chances of finding a peer to store on, it is important to choose a reasonable chunk size, as using more nodes for storage will add more overhead in finding those nodes.

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.75]{PeerStoreTrading}
  \caption{Trading protocol of PeerStore}
\end{figure}

@@

\subsection{Churn and Data Migration}
With each storage node being regular users, it is often the case that nodes can be offline. \textit{Node churn} refers to fact that peers will continuously enter and leave the P2P network \cite{StorageSearchP2PNetworks}. The designer of a P2P backup system must take this into consideration if the system is to provide reliable access to backed up data. This has been accomplished via data migration \cite{pStore,PeerStore}, where the availability of files is periodically checked and replicas are made when the nodes being used for backup are not reliably available. Other data migration algorithms have been designed specifically to work in high churn networks \cite{StorageSearchP2PNetworks}. In this case, data replicas are periodically passed to different groups of nodes, which means that newer nodes have a chance to store data. This is beneficial because data gets evenly distributed throughout the overlay network, so any given node is not more important to the network than any others with regard to the amount of data it is storing.

@@

\subsection{Cryptography}
% TODO: the figure for convergent encryption might be nice here
Since data is being stored on several untrusted peers, each user's data need to be encrypted before backup. A technique used in pStore and PeerStore is convergent encryption \cite{pStore, PeerStore}. To use this encryption mechanism, the hash of the data to encrypt is calculated, and then the hash is used as the symmetric key to encrypt the data. The main benefit of using convergent encryption is that the same data will be encrypted the same way each time, so identical blocks in the network can be shared between peers in their encrypted form.

\begin{figure}
  \centering
  \includegraphics[scale=0.75]{ConvergentEncryption}
  \caption{Convergent Encryption from PeerStore}
\end{figure}

While this is a convenient optimization, it allows for users who have access to the unencrypted version of a file to confirm whether or not that file is being stored, since no more information is needed to encrypt or decrypt a given file.
%\subsection{Case Study: pStore}
%\subsection{Case Study: PeerStore}
\section{Distributed Hash Tables}

>>>>

\begin{document}
\begin{center}
\thispagestyle{empty}
\Large{\textbf{A Major Qualifying Project Report\\ \large{ON}}}\\[0.7cm]
\LARGE{\textsc {\textbf{An Analysis of the Feasibility and Performance of BitTorrent as a Data Transfer Protocol in Peer to Peer Backup Systems}}}\\[0.5cm]
\vspace{0.2cm}
\Large{\textbf{\\Submitted to the Faculty of }}
\LARGE{\textbf{\\WORCESTER POLYTECHNIC INSTITUTE\\}}
\vspace{0.5cm}
\Large{\textbf{\\In Partial Fulfillment of the Requirement for the}}
\Large{\textbf{\\Degree of Bachelor of Science}}
\vspace{0.5cm}
\Large{\textbf{\\by}}\\[0.5cm]
\large{\textbf{Dan Bouffard}}\\
\large{\textbf{Marc Green}}\\
\vspace{0.5cm}
\large{\textbf{UNDER THE GUIDANCE OF}}\\
\large{\textbf{Professor G\'abor N. S\'ark\"ozy}}\\
\vspace{1cm}
\large{\textbf{\\May 1, 2014}}\\
\newpage
\end{center}

@@

\chapter{Background}
\section{Centralized Backup Systems}
% TODO: get reliable sources for number of users of OneDrive and Google Drive
Several popular backup systems currently in use have a centralized system design. In these systems, there are many clients that all back up their data to a central location controlled by a single entity. Common elements of these systems include a fixed amount of data storage per person, with the option to obtain more for a fee. Three of the more popular systems are Dropbox, which has over 200 million users \cite{dropboxusers}, OneDrive, with over 250 million users, and Google Drive, with over 120 million users.

@@

However, there are also several problems with using a centralized system to backup data. One of these problems is that there is a single point of failure, should the storage site be compromised in some way (for example, natural disasters). This is less of a problem for big companies with the resources to handle these situations, but there is still a cost to take the necessary preventative measures when creating the system.

@@

% TODO: reliable sources for data caps and monthly/yearly fees
Finally, there are often restrictions on what can be stored using centralized backup systems. This includes both the size of individual files and the amount of total data. More space is gained by paying a monthly or yearly fee, and even then the amount of space is capped at a specific size. For users who need to backup a large amount of large files, these systems are not very useful.

Peer to peer backup systems, on the other hand, provide an equivalent service and suffer from none of these drawbacks.

\section{Peer to Peer Backup Systems}

Peer to Peer (P2P) Backup Systems are networks that allow users to backup data without the use of a central entity. Instead, each user offers some of his or her storage space in order to use the space of others as backup locations. We start this section by introducing peer to peer networks, and then discuss the technical components of existing P2P backup systems.

@@

A peer to peer network is a distributed network model in which participants form direct connections to each other. This is in contrast to the centralized client-server network model, in which all participating clients connect to a central server to carry out their task. Client-server network models are used on the World Wide Web: the clients are users' web browsers and the server is the website being visited. In peer to peer networks, each participant, or "peer", functions as both a client and a server; peers initiating a request take on the role of the client, and peers answering the request take on the role of the server.

In the client-server model, servers are expected to be always available and accessible. Peer to peer networks do not have this luxury. Instead, they experience the constant fluctuation of peers joining and leaving the network. This is called \textit{churn}. Churn in P2P networks complicates the communication between peers because the sought peer may be offline. This is especially a problem in P2P backup systems, for example, when trying to recover data from a peer who is not online. The strategies for dealing with churn, and many other P2P-specific complications, heavily depend on the type of overlay network a P2P application uses. We discuss churn in more detail in section \ref{sec:churn}.

% TODO add reference for p2p explanation/definition

\subsubsection{Overlay Networks}

An overlay network is a network built on top of another network. That is, an overlay network consists of nodes with neighbors who are not physically connected, but instead use the underlying network to establish a virtual connection. Consider a trivial example where there are three computers connected to the Internet, but located in different parts of the world. These three computers could form an overlay network by agreeing to be virtual neighbors to each other. If their overlay network was used in a P2P application, then, when connecting to each other, they will use the underlying network, the Internet, to actually send messages. The P2P application does not need to know about the underlying network that is used to route messages, it only needs to know that there is an overlay network that defines the direct connections between peers.

As alluded to, peer to peer networks are implemented using overlay networks on top of existing infrastructure.\footnote{Overlay networks are used for many purposes, not just P2P applications. Note, however, in discussing overlay networks, we will refer to their participants as \textit{peers}.} This allows peers to form a network between each other that need not be constrained by the physical world. The responsibilities of a P2P overlay network includes providing peers the ability to join and leave the network and find other peers.

@@

We discuss structured overlay networks in more detail below, as that is the type of overlay network our P2P backup system uses.

\subsubsection{Structured Overlay Networks}

Structured overlay networks are defined by a topology and a virtual address space. For example, in our system, peers are organized in a ring and addressed by 20 byte numbers---our address space is the set of all 20 byte numbers. In other systems, peers could be organized in a 3D cube and addressed by their (x,y,z) coordinates. The topology defines the set of neighbors each peer has in the overlay. The address space serves as a way to uniquely identify peers.
% TODO maybe insert figure of chord ring topology (or our own!)

The purpose of peer to peer networks (and more broadly, networks in general) is to share data. In the P2P networks that focus not just on sharing data, but also storing, locating, and possibly retrieving data (like ours), the data shares the same virtual address space as the peers. That is, structured overlay networks give the data themselves valid addresses. For example, in our system, the data being backed up is also given a 20 byte "address". This is done to determine which peer is responsible for a given piece of data. Specifically, a peer is responsible for all data in the address space that the peer is closest to. Aside: the size of the address space for which a peer is responsible decreases as more peers join the network, which allows these types of networks to easily scale in size.

To join a structured overlay network, a peer must know at least one other peer already in the network. This peer is often called the \textit{bootstrap peer}. The joining peer will contact the bootstrap peer to find its proper place in the topology. More details about finding the proper place in the topology are given below, in the explanation of how to find other peers.

% TODO cite router.bittorrent.com as well known bootstrap node?
%In BitTorrent's structured overlay network \footnote{To clarify, we are referring to BitTorrent's distributed hash table, not its unstructured network. The latter involves using centralized trackers, the former removes such dependency. We refraim from using the term \textit{distributed hash table} to keep the explanation simple without loss of generality. We discuss distributed hash tables in section \ref{sec:dht}.}, a well known bootstrap node \footnote{The BitTorrent specification uses \textit{node} to refer to participants of its structured overlay network and \textit{peer} as a more general term \cite{bittorrentDHT}. We will adopt this convention when speaking about BitTorrent, and use them interchangeably otherwise.} is router.bittorrent.com. A BitTorrent node randomly chooses a 160 bit \textit{nodeID} to use as an address. To join the network, it will contact a bootstrap node to find the nodes it is closest to in the topology. Closeness, in BitTorrent's case, is determined by taking the $XOR$ of \textit{nodeID}s; smaller values are defined as closer \cite{bittorrentDHT}.
% TODO define XOR?
% TODO maybe remove this entire paragraph, since it doesn't really clarify anything. it just gives another example of picking an id and mysteriously finding nodes its closest to. if we end up remove it, we should reword end of previous paragraph

Leaving a structured overlay network can be as simple as not responding to any messages. Some specifications leave this functionality undefined, perhaps because peers may simply shutdown instead of gracefully quitting. Such overlay networks handle node departures as node failures. This is how our system behaves.

To find other peers, a peer in a structured overlay network will consult its routing table. This table is initially populated upon joining the network, and is kept up to date as peers join and leave. When a joining peer first contacts a bootstrap peer, it is unlikely that the bootstrap peer has the joining peer's direct neighbors in its routing table. This is because the routing table is designed to include many peers that are close, and only a few peers that are far. A joining peer, by asking the bootstrap node, will learn of the peers in the bootstrap peer's routing table that the joining node is close to. The joining peer will then contact these peers, ask for the same information, and learn of peers it is even closer to. This process repeats (usually in $O(\log n)$ time) until the joining node learns its direct neighbors \cite{p2pSurvey}.
% TODO explain O(logn)
% TODO can we relate this to our system somehow? If not, why are we talking about it? Maybe we should simplify... Though it would be unfortunate to not utilize this research into structured overlays we've done

Overlay networks are the abstraction that connects all the users of a P2P system. It is the foundation on which a P2P application can be built, and is responsible for providing methods for peers to join and leave the network, and find other peers in the network. Furthermore, structured overlay networks allow the P2P application to locate specific information very quickly because their virtual address space provides full accountability of any piece of data. It is for this reason we chose a structured overlay network as the substrate of our P2P application; it is easy to determine which peer is responsible for a given backup because we know it will be the one closest in the address space.\footnote{However, churn complicates this because a peer with an even closer address may join the network between storing and retrieving the data.}

% TODO: Ask Sarkozy/Mihaly if they think keeping this section in is worth it.
\subsection{File Chunking}
The format in which files are stored can have an effect on the efficiency of the system. At first, backing up each file as a whole might seem like the obvious choice, since files are only useful in their entirety. However, the creators of the P2P backup systems pStore and PeerStore have shown that there are  split files up into chunks, rather than distributing files to peers in their entirety. They also use special files, called File Block Lists in pStore and Descriptors in PeerStore, that list which blocks belong to the file and where to find them in the network. This has the benefit of having data units be a fixed size that will be easy to store for any node. For example, if a peer needs to store a 10GB file, it does not need to find a peer capable of storing 10GB (which might be rare); rather, it needs to find $file\_size/chunk\_size*number\_of\_replicas$ peers. Although lowering the chunk size will increase the chances of finding a peer to store on, it is important to choose a reasonable chunk size, as using more nodes for storage will add more overhead in finding those nodes.

@@

\begin{figure}
  \centering
  %\includegraphics[scale=0.75]{PeerStoreTrading}
  \caption{Trading protocol of PeerStore}
\end{figure}

@@

\subsection{Churn and Data Migration} \label{sec:churn}
With each storage node being regular users, it is often the case that nodes can be offline. \textit{Node churn} refers to fact that peers will continuously enter and leave the P2P network \cite{StorageSearchP2PNetworks}. The designer of a P2P backup system must take this into consideration if the system is to provide reliable access to backed up data. This has been accomplished via data migration \cite{pStore,PeerStore}, where the availability of files is periodically checked and replicas are made when the nodes being used for backup are not reliably available. Other data migration algorithms have been designed specifically to work in high churn networks \cite{StorageSearchP2PNetworks}. In this case, data replicas are periodically passed to different groups of nodes, which means that newer nodes have a chance to store data. This is beneficial because data gets evenly distributed throughout the overlay network, so any given node is not more important to the network than any others with regard to the amount of data it is storing.

@@

\subsection{Cryptography} \label{sec:crypto}
% TODO: the figure for convergent encryption might be nice here
Since data is being stored on several untrusted peers, each user's data need to be encrypted before backup. A technique used in pStore and PeerStore is convergent encryption \cite{pStore, PeerStore}. To use this encryption mechanism, the hash of the data to encrypt is calculated, and then the hash is used as the symmetric key to encrypt the data. The main benefit of using convergent encryption is that the same data will be encrypted the same way each time, so identical blocks in the network can be shared between peers in their encrypted form.

\begin{figure}
  \centering
  %\includegraphics[scale=0.75]{ConvergentEncryption}
  \caption{Convergent Encryption from PeerStore}
\end{figure}

While this is a convenient optimization, it allows for users who have access to the unencrypted version of a file to confirm whether or not that file is being stored, since no more information is needed to encrypt or decrypt a given file.
%\subsection{Case Study: pStore}
%\subsection{Case Study: PeerStore}
\section{Distributed Hash Tables} \label{sec:dht}
